Targeted interrogation of mutational hotspots in rheumatoid arthritis

### 2024-03-11 scomatic/reports/targeted_mutation_calling.Rmd
extend windows for targeted_mutation_calling, fix plot dims

# libraries
library(magrittr)
library(ggplot2)
library(GenomicRanges)

@@

# reads
reads <-
  list.files(pattern = '_reads.tsv') %>%
  purrr::set_names(., gsub('_reads.tsv', '', .)) %>%
  purrr::map(function(file) {
    readr::read_tsv(file, show_col_types = F)
  }) %>%
  dplyr::bind_rows(.id = 'id') %>%
  tidyr::pivot_longer(cols = dplyr::all_of(nucleotides)) %>%
  dplyr::mutate(
    # colour in alt alleles
    status = dplyr::case_when(toupper(name) == ref ~ 'ref',
                              name %in% c('-', '_') ~ 'DEL',
                              TRUE ~ toupper(name)) %>%
      forcats::fct_relevel(names(nt_cols))) %>%
  # add detected mutations
  dplyr::left_join(recurmuts, by = c('chr', 'mut_pos')) %>%
  split(., .$gene)
```

@@

```{r plot}
# function: create plot list
plot_positions <- function(reads, include_indels = T) {
  p <-
    names(reads) %>%
    purrr::set_names() %>%
    purrr::map(function(gene) {

      p_dat <-
        reads[[gene]] %>%
        # optionally exclude indels
        dplyr::filter(include_indels == T |
                !name %in% c('-', '_', 'INS', 'ins')) %>%
        # only plot celltypes with any reads
        dplyr::group_by(celltype) %>%
        dplyr::filter(max(value) > 0) %>%
        dplyr::ungroup()

@@

        g_title <- paste0(
          gene, ' w/', 
          ifelse(include_indels == T, '', 'o'), 
          ' indels')

        # get plot dims
        g_p_dims <-
          c(min(max(5, dplyr::n_distinct(p_dat$mut_pos) * 2 + 2), 50),
            min(max(5, dplyr::n_distinct(p_dat$celltype) * 1.7 + 2), 50))

@@

# generate plot lists
p_w_indels <- plot_positions(reads)
p_wo_indels <- plot_positions(reads, F)
```

@@

  knit_plot <- c(
    '',
    paste0('```{r ', gene,'_w_indels, fig.dim = p_w_indels[["', gene, '"]]$dims, warning = F, message = F, echo = F}'),
    paste0('print(p_w_indels[["', gene, '"]]$plot)'),
    '```',
    '',
    paste0('```{r ', gene,'_wo_indels, fig.dim = p_wo_indels[["', gene, '"]]$dims, warning = F, message = F, echo = F}'),
    paste0('print(p_wo_indels[["', gene, '"]]$plot)'),
    '```',
    ''
  )

@@

## Mutations types, by patient across genes

```{r muts_by_patient, fig.width = 10}
reads %>%
  dplyr::bind_rows() %>%
  ggplot(aes(x = id, y = value, fill = status)) +
  geom_col() +
  scale_fill_manual(values = nt_cols) + 
  theme_classic() +
  scale_x_discrete(guide = guide_axis(angle = -90))
```

>>>>

# libraries
library(magrittr)
library(ggplot2)

@@

# reads
reads <-
  list.files(pattern = '_reads.tsv') %>%
  purrr::set_names(., gsub('_reads.tsv', '', .)) %>%
  purrr::map(function(file) {
    readr::read_tsv(file, show_col_types = F)
  }) %>%
  dplyr::bind_rows(.id = 'id') %>%
  tidyr::pivot_longer(cols = dplyr::all_of(nucleotides)) %>%
  dplyr::mutate(
    # colour in alt alleles
    status = dplyr::case_when(toupper(name) == ref ~ 'ref',
                              name %in% c('-', '_') ~ 'DEL',
                              TRUE ~ toupper(name)) %>%
      forcats::fct_relevel(names(nt_cols))) %>%
  # add detected mutations
  dplyr::left_join(recurmuts, by = c('chr', 'mut_pos')) %>%
  split(., .$gene)

# save reads
reads %>% 
  saveRDS('reads.rds')
```

@@

```{r plot}
# function: create plot list
plot_positions <- function(reads, include_indels = T, include_refs = T) {
  p <-
    names(reads) %>%
    purrr::set_names() %>%
    purrr::map(function(gene) {

      p_dat <-
        reads[[gene]] %>%
        # optionally exclude indels
        dplyr::filter(
          include_indels == T | !status %in% c('DEL', 'INS'),
          include_refs == T | status != 'ref') %>%
        # only plot celltypes with any reads
        dplyr::group_by(celltype) %>%
        dplyr::filter(max(value) > 0) %>%
        dplyr::ungroup()

@@

        g_title <- paste0(
          gene, ', w/', 
          ifelse(include_indels == T, '', 'o'), 
          ' indels',
          ifelse(include_refs == T, '', ', w/o refs'))

        # get plot dims
        g_p_dims <-
          c(min(max(7, (dplyr::n_distinct(p_dat$mut_pos) + 2) * 2), 50),
            min(max(7, (dplyr::n_distinct(p_dat$celltype)) * 1.5), 50))

@@

# generate plot lists
p_w_indels <- plot_positions(reads)
p_wo_indels <- plot_positions(reads, F)
p_wo_indels_wo_refs <- plot_positions(reads, F, F)
```

@@

  knit_plot <- c(
    '',
    paste0('```{r ', gene,'_w_indels, fig.dim = p_w_indels[["', gene, '"]]$dims, warning = F, message = F, echo = F}'),
    paste0('print(p_w_indels[["', gene, '"]]$plot)'),
    '```',
    '',
    paste0('```{r ', gene,'_wo_indels, fig.dim = p_wo_indels[["', gene, '"]]$dims, warning = F, message = F, echo = F}'),
    paste0('print(p_wo_indels[["', gene, '"]]$plot)'),
    '```',
    '',
    paste0('```{r ', gene,'_wo_indels_wo_refs, fig.dim = p_wo_indels_wo_refs[["', gene, '"]]$dims, warning = F, message = F, echo = F}'),
    paste0('print(p_wo_indels_wo_refs[["', gene, '"]]$plot)'),
    '```',
    ''
  )

@@

## Mutation types, by patient across genes

### Mutation types, by patients across all genes

```{r muts_by_patient, fig.width = 10, class.source = 'fold-hide'}
reads %>%
  dplyr::bind_rows() %>%
  ggplot(aes(x = id, y = value, fill = status)) +
  geom_col() +
  scale_fill_manual(values = nt_cols) + 
  theme_classic() +
  scale_x_discrete(guide = guide_axis(angle = -90))
```

### Mutation types, by patient across all genes, without indels

```{r muts_by_patient_no_indels, fig.width = 10, class.source = 'fold-hide'}
reads %>%
  dplyr::bind_rows() %>%
  dplyr::filter(!status %in% c('INS', 'DEL', 'ref')) %>%
  ggplot(aes(x = id, y = value, fill = status)) +
  geom_col() +
  scale_fill_manual(values = nt_cols) + 
  theme_classic() +
  scale_x_discrete(guide = guide_axis(angle = -90))
```
